Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-11 docs/proposals/NSF/ProjDesc/ProjDesc.tex
NM

%---------------------------------------------------------------------------------------------------%
%% PACKAGES
\usepackage{fullpage}
%\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{graphicx, amsmath, amssymb, amsfonts, mathtools, mathrsfs, color}
\usepackage{comment, enumerate, tabularx}
\usepackage{natbib, hyperref, url}
%\usepackage[justification=RaggedRight]{caption}
\usepackage{outlines,enumitem}
\usepackage{wrapfig}
%---------------------------------------------------------------------------------------------------%

@@

Naturally occurring porous and granular materials, such as soil, sand, and clay, play a pivotal role in regulating earth's water resources by filtering contaminants and, over long timescales, supplying fresh water. Not only is this process essential for human water resources, but also for the ecology of rivers, estuaries, and other natural habitats. Natural water resources, however, have been placed under enormous pressure by human population growth and associated activities that can comprise the natural filtration cycle, for example expansion of industry, urbanization, pollution, and climate change. See the USGS and the UN report on water resources for further detail on the threats to water resources [USGS?] \cite{UNwater}.

To take one example of particular importance to the state of Florida, the formation of sinkholes can be triggered by certain human activities such as groundwater pumping or the weight of run-off storage ponds \cite{sandhu2018fate}. When collapse occurs below a storage pond, the underlying reservoir of pristine groundwater can be severely contaminated, thus surrendering a natural resource of great economic and societal value. Clearly, a deeper understanding of the mechanisms leading to sinkhole formation, and more generally that underly earth's natural filtration system, could reap enormous benefits in predicting and mitigating such environmental disasters. Other applications include understanding and predicting contamination by nuclear waste and by hydraulic fracturing.

This proposal aims to develop a deeper understanding of the physical processes by which groundwater flow can alter properties of a porous medium. A recognized complexity of naturally occurring media is that their macroscopic properties, e.g.~permeability and mechanical strength, can be highly heterogeneous and anisotropic {\color{blue}(cite)}. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely. The main objective of this proposal is to gain a deeper understanding of such processes through analyzing a specific set of problems and by developing a host of new mathematical tools, ranging from analytical to computational ones.

Physical mechanisms that can dynamically alter porous-medium properties include:
\begin{itemize}
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of grains due to intervening fluid flow and due to sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse. 
In particular, we view the slow processes of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of grains. This modified distribution can in turn affect transport by enabling smaller grains to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of grains. For instance, erosion can carve grains into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. {\color{blue} Need to include comment on dissolution changing conduit paths.}

@@

Only recently have these ideas been extended to low Reynolds number flows  \cite{mitchell2017generalized, Quaife2018} --- the regime applicable to porous media as well as biological applications like plaque erosion. While the governing fluid-flow equations become linear in this regime, porous-media applications present a host of new challenges, for example:
\begin{itemize}
\item Porous media typically involves dense packings so that close-range effects and even nearly contacting bodies must be considered {\color{blue}(cite nearly contact papers)}.
\item The Stokes limit introduces {\em longer} range effects than the inertial limit does, thus coupling the shape evolution of {\em all} grains and conduits to a higher degree.
\item Somewhat counterintuitively, erosion does not always act to smooth shapes but rather can create sharp features and corners. These features can be difficult to resolve and, further, can trigger numerical instabilities. Dissolution, meanwhile, can initiate a roughening transition {\color{blue} (cite)}, leading to similar numerical difficulties.
\item Additionally effects, such as sedimentation, transport, and compaction constantly work to reconfigure the medium as its constituents and boundaries degrade.
\end{itemize}
These challenges require innovative computational techniques, many of which have been developed recently \cite{Quaife2018} and many of which await resolution. The grand goal of achieving high-fidelity simulations of order-100 mobile bodies which undergo simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency to run large ensembles for statistical analysis, still awaits. Fortunately, some closely related laboratory experiments on erosion in colloidal systems are already being explored in the Datta Lab at Princeton \cite{bizmark2019multiscale}. The experiments offer a wonderful opportunity to guide, calibrate, and validate theoretical developments.

While erosion and dissolution correspond to gradual changes in a medium, the example of a sinkhole represents a sudden, catastrophic event. It is likely that all of the mechanisms listed above --- erosion, dissolution, transport, compaction, and gravitational collapse --- contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity. Sinkhole formation has been explored with preliminary experiments run at the Geophysical Fluid Dynamics Institute at Florida State University \cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers.

	With two Mathematics PhD students, the PI has begun running preliminary simulations to model this behaviors. The simulations, which are at an early stage of development, are based on a hybrid continuum-discrete model for the permeability field and grain particle field respectively. They utilize a homogenized description of the flow field, e.g.~the Darcy or Brinkman equations.
These simulations incorporate the physical effects of:
\begin{itemize}
\item Transport of grains due to seepage force from the surrounding fluid flow.
\item Gravitational sedimentation of grains.
\item Feedback of grain-distribution onto permeability field.
\item Contact and cohesive forces between grains.
\end{itemize}
Preliminary investigation suggests that this last mechanisms of cohesion is essential to model sinkhole collapse in the system, and the simulations await careful calibration against experimental data.

@@

% McCurdy and Wang work
% Multiphase stuff too

% Could emphasize dissolution of karst more too.
% Not crazy about inverse model sentences, maybe move to end of paragraph.

%I could try to work in more mathy stuff: complex dynamical system, feedback, timescales, computational methods, analytical reductions...

>>>>

%---------------------------------------------------------------------------------------------------%
%% PACKAGES
\usepackage{fullpage}
%\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{graphicx, amsmath, amssymb, amsfonts, mathtools, mathrsfs, color}
\usepackage{comment, enumerate, tabularx}
\usepackage{natbib, hyperref, url}
%\usepackage[justification=RaggedRight]{caption}
\usepackage{outlines,enumitem}
\usepackage{wrapfig, todonotes}
%---------------------------------------------------------------------------------------------------%

@@

	Naturally occurring porous and granular materials, such as soil, sand, and clay, play a pivotal role in regulating earth's water resources by filtering contaminants and, over long timescales, supplying fresh water. Not only is this process essential for human water resources, but also for the ecology of rivers, estuaries, and other natural habitats. Natural water resources, however, have been placed under enormous pressure by human population growth and associated activities that can comprise the natural filtration cycle, for example expansion of industry, urbanization, pollution, and climate change. See the UN report on water resources for further detail on the threats to water resources \cite{UNwater}.
% Could also include USGS if needed.

	These filtration processes, however, rely on an understand of complex transport phenomenon in porous media. Further, the action of continually flowing groundwater can alter medium properties through erosive and transportive processes. Thus, it becomes necessary to understand the transport of tracers through complex media whose properties are changing {\em dynamically} in response to the intervening fluid flow.

	This proposal aims to develop a deeper understanding of the physical processes by which groundwater flow can alter properties of a porous medium and to characterize the associated changes in transport, e.g.~anomalous dispersion. A recognized complexity of naturally occurring media is that their macroscopic properties, e.g.~permeability and mechanical strength, can be highly heterogeneous and anisotropic. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.

Physical mechanisms that can dynamically alter porous-medium properties include:
\begin{itemize}
\item Erosion of solid material due to fluid-mechanical stresses or chemical dissolution.
\item Transport of grains due to intervening fluid flow and due to sedimentation.
\item Compaction of the medium due to overlying loads.
\item Gravitational collapse of a medium due to loads exceeding a threshold.
\end{itemize}
The above are roughly ordered according to timescale, from the slowest acting to the fastest. While many of these mechanisms have been studied individually in various contexts and to varying levels of rigor, very little is known about their combined effect on the dynamic evolution of a porous medium. Since we aim to build a quantitative model from the ground up, we will first focus on the slowest acting mechanisms, as these ultimately set the stage for the more rapid events like sinkhole collapse.

In particular, we view the slow processes of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of grains. This modified distribution can in turn affect transport by enabling smaller grains to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of grains. For instance, erosion can carve grains into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties \cite{szymczak2009wormhole}.

@@

Only recently have these ideas been extended to low Reynolds number flows  \cite{mitchell2017generalized, Quaife2018} --- the regime applicable to porous media. While the governing fluid-flow equations become linear in this regime, porous-media applications present a host of new challenges, for example:
\begin{itemize}
\item Porous media typically involves dense packings so that close-range effects and even nearly contacting bodies must be considered.
\item The Stokes limit introduces {\em longer} range effects than the inertial limit does, thus coupling the shape evolution of {\em all} grains and conduits to a higher degree.
\item Somewhat counterintuitively, erosion does not always act to smooth shapes but rather can create sharp features and corners \cite{Ristroph2012}. These features can be difficult to resolve and, further, can trigger numerical instabilities \cite{Quaife2018}. Dissolution, meanwhile, can initiate a roughening transition \cite{claudin2017dissolution}, leading to similar numerical difficulties.
\item Additionally effects, such as sedimentation, transport, and compaction constantly work to reconfigure the medium as its constituents and boundaries degrade.
\end{itemize}
These challenges require innovative computational techniques, many of which have been developed recently \cite{Quaife2018} and many of which await resolution. An ultimate goal of this proposal is to achieve high-fidelity simulations of $O(100-1000)$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency to run large ensembles for statistical analysis.
Fortunately, some closely related laboratory experiments on erosion in colloidal systems are already being explored in the Datta Lab at Princeton \cite{bizmark2019multiscale}. The experiments offer a wonderful opportunity to guide, calibrate, and validate theoretical developments.

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse can occur suddenly and catastrophically. It is likely that all of the mechanisms listed above --- erosion, dissolution, transport, compaction, and gravitational collapse --- contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity. Sinkhole formation has been explored with preliminary experiments run at the Geophysical Fluid Dynamics Institute at Florida State University \cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers.

\todo[inline]{I am not sure if we will keep this much detail in the paragraphs below on the sinkhole stuff. It depends on how much weight we give sinkholes in the body of the proposal. Instead, we could replace this discussion with more depth about the tracer transport and anomalous dispersion.}

	With two Mathematics PhD students, PI Moore has begun running preliminary simulations to model this behaviors. The simulations, which are at an early stage of development, are based on a hybrid continuum-discrete model for the permeability field and grain particle field respectively. They utilize a homogenized description of the flow field, e.g.~the Darcy or Brinkman equations.
These simulations incorporate the physical effects of:
\begin{itemize}
\item Transport of grains due to seepage force from the surrounding fluid flow.
\item Gravitational sedimentation of grains.
\item Feedback of grain-distribution onto permeability field.
\item Contact and cohesive forces between grains.
\end{itemize}
Preliminary investigation suggests that this last mechanisms of cohesion is essential to model sinkhole collapse in the system, and the simulations await careful calibration against experimental data.

@@

% Stuff we could include
% McCurdy and Wang work
% Multiphase work

% Removed this paragraph because it emphasized sinkholes too much.
%To take one example (of particular relevance in the state of Florida) the formation of sinkholes can be triggered by certain human activities such as groundwater pumping or the weight of run-off storage ponds \cite{sandhu2018fate}. When collapse occurs below a storage pond, the underlying reservoir of pristine groundwater can be severely contaminated, thus surrendering a natural resource of great economic and societal value. Clearly, a deeper understanding of the mechanisms leading to sinkhole formation, and more generally that underly earth's natural filtration system, could reap enormous benefits in predicting and mitigating such environmental disasters. Other applications include understanding and predicting contamination by nuclear waste and by hydraulic fracturing.
